---
title: "EDA"
author: "Ariel Wentworth"
date: "September 2020"
output: pdf_document
---

```{r libraries}
library('tidyverse')
```


# The Gender Gap: by Degree Level
We will first do some exploration of degrees conferred, separated by degree level.

## Doctor's
```{r read doctors}
doctors <- read_csv('data/doctors.csv')
head(doctors)
```

```{r select: doctors by sex}
doctors_by_sex <- doctors %>%
  select(Sex, Year, Total) %>%
  group_by(Year) %>%
  filter(Sex!="Total")

head(doctors_by_sex)
```

```{r grouped eda doctors}
ggplot(doctors_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("U.S. doctoral degrees conferred over time, by gender") +
  geom_bar(position="dodge", stat="identity") +
  theme(axis.text.x = element_text(angle = 90))
```

Though females seem to be steadily increasing over time, male degrees fluctate much more.

```{r stacked eda doctors}
ggplot(doctors_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("U.S. doctoral degrees conferred over time, by gender") +
  geom_bar(position="stack",stat="identity") +
  theme(axis.text.x = element_text(angle = 90))
```

```{r proportion eda doctors}
ggplot(doctors_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("Proportion of U.S. doctoral degrees conferred over time, by gender") +
  geom_bar(position="fill", stat="identity") +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_abline(slope=0, intercept = 0.5)
```

Around 2005 was when females and males were earning doctorate degrees equally. Since then, females have been taking an increasing amount of the proportion, but still not startling more than 50%.

## Masters
```{r read masters}
masters <- read_csv('data/masters.csv')
head(masters)
```

```{r select: masters by sex}
masters_by_sex <- masters %>%
  select(Sex, Year, Total) %>%
  group_by(Year) %>%
  filter(Sex!="Total")

head(masters_by_sex)
```

```{r grouped eda masters}
ggplot(masters_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("U.S. masters degrees conferred over time, by gender") +
  geom_bar(position="dodge", stat="identity") +
  theme(axis.text.x = element_text(angle = 90))
```

Though both male and female masters degrees seem to be gaining over time, there has been increasingly more degrees conferred to females, especially in the 21st century.

```{r stacked eda masters}
ggplot(masters_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("U.S. masters degrees conferred over time, by gender") +
  geom_bar(position="stack",stat="identity") +
  theme(axis.text.x = element_text(angle = 90))
```

```{r proportion eda masters}
ggplot(masters_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("Proportion of U.S. masters degrees conferred over time, by gender") +
  geom_bar(position="fill", stat="identity") +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_abline(slope=0, intercept = 0.5)
```

Around 1981, women and men were earning bachelors degrees at similar rates, and into the 21st century females starting gaining on males much more rapidly. Since the start of the 21st century, only incremental changes have occured.

## Bachelors
```{r read bachlors}
bachelors <- read_csv('data/bachelors.csv')
head(bachelors)
```

```{r select: bachelors by sex}
bachelors_by_sex <- bachelors %>%
  select(Sex, Year, Total) %>%
  group_by(Year) %>%
  filter(Sex!="Total")

head(bachelors_by_sex)
```

```{r grouped eda bachelors}
ggplot(bachelors_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("U.S. bachelors degrees conferred over time, by gender") +
  geom_bar(position="dodge", stat="identity") +
  theme(axis.text.x = element_text(angle = 90))
```

Both males and females are increasing over time, though it seems females may be increasing more rapidly.

```{r stacked eda bachelors}
ggplot(bachelors_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("U.S. bachlors degrees conferred over time, by gender") +
  geom_bar(position="stack",stat="identity") +
  theme(axis.text.x = element_text(angle = 90))
```

```{r proportion eda bachelors}
ggplot(bachelors_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("Proportion of U.S. bachelors degrees conferred over time, by gender") +
  geom_bar(position="fill", stat="identity") +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_abline(slope=0, intercept = 0.5)
```

Around 1981, women and men were earning equal amounts of bachelor's degrees. Since about 2000, it seems that the proportion of women vs men earning bachelor's degrees has stayed pretty consistent around a 60-40 split.

## Associates
```{r read associates}
associates <- read_csv('data/associates.csv')
head(associates)
```

```{r select: associates by sex}
associates_by_sex <- associates %>%
  select(Sex, Year, Total) %>%
  group_by(Year) %>%
  filter(Sex!="Total")

head(associates_by_sex)
```

```{r grouped eda associates}
ggplot(associates_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("U.S. associates degrees conferred over time, by gender") +
  geom_bar(position="dodge", stat="identity") +
  theme(axis.text.x = element_text(angle = 90))
```

Both males and females are mostly increasing over time (potentially with a peak around 2011), though it seems females are increasing much more rapidly.

```{r stacked eda associates}
ggplot(associates_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("U.S. associates degrees conferred over time, by gender") +
  geom_bar(position="stack",stat="identity") +
  theme(axis.text.x = element_text(angle = 90))
```

```{r proportion eda associates}
ggplot(associates_by_sex, aes(fill=Sex, y=Total, x=Year)) + 
  ggtitle("Proportion of U.S. associates degrees conferred over time, by gender") +
  geom_bar(position="fill", stat="identity") +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_abline(slope=0, intercept = 0.5)
```

Since the 80's women have been earning more associate's degrees than men. The difference here is much more pronouced than in higher levels of education, and is extremely persistent today.

